---
title: Imani STA9750 
---
My name is Imani. I'm from New Orleans, La. I moved to upstate New York when I joined the military. I obtained my undergraduate degree at CUNY York College and now I will obtain my graduate degree at Baruch - Zicklin School of Business in Business Analytics. I love to travel and try new foods.
```{r}
#| include: false
1+1
```
--------------
Last Updated: `r format(Sys.time(), "%A %m %d, %Y at %H:%M%p")`
```{r}
#| echo: false
#| message: false
#| warning: false

if(!require("leaflet")){
    options(repos=c(CRAN="https://cloud.r-project.org"))
    install.packages("leaflet")
    stopifnot(require("leaflet"))
}

baruch_longitude <- -73.98333
baruch_latitude  <- +40.75028

leaflet() |>
  addTiles() |>
  setView(baruch_longitude, baruch_latitude, zoom=17) |>
  addPopups(baruch_longitude, baruch_latitude, 
            "I am a Master's student at <b>Baruch College</b>!")
```
#!/usr/bin/env Rscript
if(!require("quarto")){
    install.packages("quarto")
}
library(quarto)
if(!quarto::quarto_binary_sitrep()){
    stop("Something is wrong with your quarto installation.")
}
quarto::quarto_render(".")
system("git add docs/*")
if(!any(grepl("rstudio", search()))){q("no")}
